---
title: "CSI 873, EM Homework"
author: "Thomas W. Jones"
date: "10/25/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Overview

As instructed, I implement the EM algorithm for clustering with two cluster centers. The key assumptions are that the data are generated by drawing from two Gaussian distributions with standard deviations of 1. The parameters to be estimated are the means of these Gaussian distributions.

The algorithm converged after 18 iterations. The two means discovered were $\mu_1 = 0.399$ and $\mu_2 = -0.647$. Figure 1, below, is a histogram of the data with normal overlays of the two learned distributions. Figure 2, below, plots the sum of absolute differences between learned means at each interation.

```{r plots, echo = FALSE}
load("hwk_em.RData")

# Histogram of clustering result with normal overlays
hist(d, breaks = 100, col = rgb(0,0,0,0.4), freq = FALSE,
     main = "EM Clustering Result",
     xlab = "Range of Data")
abline(v = result$centers, lwd = 2, col = c("blue", "red"))
lines(seq(-5,5,by = 0.05), 
      dnorm(seq(-5,5,by = 0.05), mean = result$centers[ 1 ], sd = 1), 
      col = "blue", lty = 2, lwd = 2)
lines(seq(-5,5,by = 0.05), 
      dnorm(seq(-5,5,by = 0.05), mean = result$centers[ 2 ], sd = 1), 
      col = "red", lty = 2, lwd = 2)
legend("topright", 
       legend = c(expression(paste(mu[ 1 ], "= 0.399")),
                  expression(paste(mu[ 2 ], "= -0.647"))),
       col = c("blue", "red"), lwd = 2)

# Plot of cost history for each iteration
plot(result$cost_history, type = "o", lwd = 3, pch = 19,
     col = "#66c2a5",
     main = "Sum of Absolute Differences \nin Means Between Iterations", 
     xlab = "Number of Iterations",
     ylab = "")
```


### Modeling approach

In my approach, I made decisions on starting points and convergence criteria. I chose to initialize $\mu_1$ and $\mu_2$ non-randomly. I partition the data in two halves and take the means of each half as initial values. I chose to stop iteration when the sum of absolute differences in means between runs were less than $0.0001$, or $10^{-4}$. Symbolically, 

$|\mu_1^{(t)} - \mu_1^{(t-1)}| + |\mu_1^{(t)} - \mu_1^{(t-1)}| < 10 ^ {-4}$.

### Instructions on running the code
The code below should be straightforward to use. It relys on no external packages. I wrote and tested the below code using R version 3.4.2 on macOS Sierra 10.12.6. It should run just fine on most operating systems, including Windows. I recommend using the RStudio IDE, but it is not necessary. I am happy to provide more specific instructions if I know your operating system.

To run the code, do the following:

1. Create a folder called `data_raw` as a subfolder of your working directory.
2. Unzip the data files you sent us inside of `data_raw`.
3. Open an R console or RStudio console.
4. If necessary, switch to your working directory using the `setwd` function.
5. Copy and paste the below code into the R console and press "enter".

### The code (itself)

```{r code, eval = FALSE}


rm(list = ls())

### Load and prepare the data --------------------------------------------------
d <- scan("data_raw/em/Jones.txt", what = "", sep = "\n")
d <- unlist(strsplit(d, " "))
d <- as.numeric(d)

### Run EM Algorithm -----------------------------------------------------------

# initialize random cluster centers by taking means of the largest/smallest
# halves of the data

mu_1 <- mean(sort(d, decreasing = TRUE)[ 1:(length(d) / 2) ])
mu_2 <- mean(sort(d)[ 1:length(d) / 2 ])

# initialize some variables used for looping

mu_diff <- 100

cost_history <- c()

step <- 0

# loop until convergence

while (mu_diff > 0.0001) {
  
  step <- step + 1
  
  # e step
  e_1 <- dnorm(d, mean = mu_1, sd = 1) / 
    (dnorm(d, mean = mu_1, sd = 1) + dnorm(d, mean = mu_2, sd = 1))
  
  e_2 <- dnorm(d, mean = mu_2, sd = 1) / 
    (dnorm(d, mean = mu_1, sd = 1) + dnorm(d, mean = mu_2, sd = 1))
  
  # update mu_last for comparison to assess convergence
  mu_last <- c(mu_1, mu_2)
  
  # m step
  mu_1 <- as.numeric((t(d) %*% e_1) / sum(e_1))
  
  mu_2 <- as.numeric((t(d) %*% e_2) / sum(e_2))
  
  # update mu_diff as sum of absolute differences in means between iterations
  mu_diff <- sum(abs(c(mu_1, mu_2) - mu_last))
  
  cost_history <- c(cost_history, mu_diff)
  
}

# Once converged, store result as a single object
result <- list(centers = c(mu_1 = mu_1, mu_2 = mu_2), # found means
               assignments = data.frame(d = d, # probabilistic assignments
                                        e_1 = e_1,
                                        e_2 = e_2,
                                        stringsAsFactors = FALSE),
               cost_history = cost_history) # convergence by iterations

# Save result
save(d, result, file = "hwk_em.RData")

# Histogram of clustering result with normal overlays
hist(d, breaks = 100, col = rgb(0,0,0,0.4), freq = FALSE,
     main = "EM Clustering Result",
     xlab = "Range of Data")
abline(v = result$centers, lwd = 2, col = c("blue", "red"))
lines(seq(-5,5,by = 0.05), 
      dnorm(seq(-5,5,by = 0.05), mean = result$centers[ 1 ], sd = 1), 
      col = "blue", lty = 2, lwd = 2)
lines(seq(-5,5,by = 0.05), 
      dnorm(seq(-5,5,by = 0.05), mean = result$centers[ 2 ], sd = 1), 
      col = "red", lty = 2, lwd = 2)
legend("topright", 
       legend = c(expression(paste(mu[ 1 ], "= 0.399")),
                  expression(paste(mu[ 2 ], "= -0.647"))),
       col = c("blue", "red"), lwd = 2)

# Plot of cost history for each iteration
plot(result$cost_history, type = "l", lwd = 3,
     col = "#66c2a5",
     main = "Sum of Absolute Differences \nin Means Between Iterations", 
     xlab = "Number of Iterations",
     ylab = "")


```

